---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
1. Unzip folder and read enclosed .csv file into a data frame
2. Convert data frame to a tibble
```{r load, results="hide"}
files <- unzip("activity.zip", list = T)
data <- read.csv(unz("activity.zip", files$Name[1]))
library(tidyverse)
data <- as_tibble(data)
```

## What is mean total number of steps taken per day?
Remove missing values and calculate step totals for each day
```{r step totals}
data$date <- as.Date(data$date, "%Y-%m-%d")
step_totals <- data %>% 
        drop_na() %>% 
        group_by(date) %>% 
        summarize(steps = sum(steps))

```
Make a histogram of number of steps per day
``` {r stephist}
p <- qplot(step_totals$steps)
p + stat_bin(binwidth = 2000) + 
  labs(title = "Frequency of total number of steps taken per day",
           x = "Total steps taken per day", y = "Frequency (days)")
```

Calculate and report the mean and median of total steps per day
```{r stepdailyaverages, results='asis'}
daily_step_summary <- step_totals %>% 
  summarize(mean = mean(steps), median = median(steps))
library(xtable)
xt <- xtable(x = daily_step_summary, caption = "Average steps per day")
print(xt, "html")
```

## What is the average daily activity pattern?
- Get average steps for each interval across all days
- Plot as a time-series plot
- Calculate and report interval with max average steps across all days
``` {r stepsbyinterval}
interval_avg <- data %>%
  drop_na() %>% 
  group_by(interval) %>% 
  summarize(avg = mean(steps))


plot(x = interval_avg$interval, y = interval_avg$avg, type = "l", 
     xlab = "Interval (minutes after 12AM)", 
     ylab = "Steps taken during 5 min interval",
     main = "Average steps taken during 5-minute intervals across multiple days")

interval_avg %>% 
  arrange(desc(avg)) %>% 
  slice(1)
```

## Imputing missing values
Calculate number of missing values
``` {r missingvalues}
data %>% summarize(missing.values = sum(is.na(.)))
```

- Fill in missing values with mean for that interval across all days
``` {r fillmissing}
data_full <- left_join(data, interval_avg)
data_full <- data_full %>% 
  mutate(steps = coalesce(steps, avg))
```  

- Make a histogram of total steps per day
```{r histimputed}
imputed_step_totals <- data_full %>% 
        group_by(date) %>% 
        summarize(steps = sum(steps))

p <- qplot(imputed_step_totals$steps)
p + stat_bin(binwidth = 2000) + 
  labs(title = "Frequency of total number of steps taken per day
       (missing values removed)",
           x = "Total steps taken per day", y = "Frequency (days)")
```  

- Calculate and report mean and median total steps per day  
```{r imputedaverages, results='asis'}
imputed_daily_step_summary <- imputed_step_totals %>% 
  summarize(mean = mean(steps), median = median(steps))
library(xtable)
xt <- xtable(x = imputed_daily_step_summary, caption = "Average steps per day
             (missing values removed)")
print(xt, "html")

```
*This method of imputing missing data (using the average for that interval
across all days) did not impact the mean and median total number of steps
taken per day.*  

## Are there differences in activity patterns between weekdays and weekends?
Add weekday factor to imputed data set
``` {r weekday}
library(lubridate)
data_full <- data_full %>% 
  mutate(weekdays = wday(date, abbr = T)) %>% 
  mutate(weekdays = if_else(weekdays < 7 & weekdays > 1, "Weekday", "Weekend")) %>% 
  mutate(weekdays = as.factor(weekdays))
```
Make a panel plot comparing average number of steps during 5-minute intervals
throughout the day across weekdays versus weekends
``` {r weekdaysteps}
wd_interval_avg <- data_full %>%
  group_by(interval, weekdays) %>% 
  summarize(avg = mean(steps))

p <- ggplot(data = wd_interval_avg)
p + geom_line(mapping = aes(x = interval, y = avg)) + 
  facet_wrap(~ weekdays, nrow = 2) + labs(
  x= "Interval (minutes after 12AM)", 
  y = "Steps taken during 5 min interval",
  title = "Average steps taken during 5-minute intervals across multiple days"
)

```